---
title: "Defining 'rivers' for the galamsey study — OSM vs MERIT Hydro vs remote sensing"
subtitle: "An exploration on the Ankobra basin — notes and open questions"
author: "Till Meissner"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6, dpi = 110)
pacman::p_load(terra, sf, leaflet, tidyverse, here, janitor, scales,
               patchwork, htmltools, conflicted)
conflicts_prefer(dplyr::filter, dplyr::select, dplyr::mutate,
                 dplyr::summarise, base::intersect, base::union)
UTM30N <- 32630

# Bright, high-contrast palette for the leaflet layers (they sit on a dark
# satellite basemap, where muted blues/reds/greys wash out).
col_osm_nat <- "#3DE1FF"  # cyan  — OSM natural watercourses
col_osm_art <- "#FF7A5C"  # coral — OSM artificial channels
col_osm_oth <- "#FFE14D"  # yellow— OSM other/review
col_upa10   <- "#FF9500"  # orange— MERIT channels, low cut
col_upa50   <- "#FF3DDA"  # magenta MERIT channels, high cut
col_gal     <- "#39FF14"  # neon green — galamsey polygons

# Native-4326 MERIT Ankobra clip (bands: dir, upa, wth, elv). See d_04_merit.R.
merit_path      <- here("data", "raw", "merit", "merit_hydro_ankobra.tif")
osm_path        <- here("data", "raw", "shapefiles", "osm_waterways", "waterways_lines.shp")
barenblitt_path <- here("data", "raw", "barenblitt", "FullConversiontoMiningExtent2019.shp")
hex_cache_path  <- here("data", "processed", "hex_5km_crosssection.rds")
```

# What this research project is about {#background}

*(Context for a reader coming to this cold — feel free to skip to [the specific question](#intro).)*

**Galamsey** is small-scale, largely unlicensed **artisanal gold mining**, which has expanded rapidly
across southern Ghana over the last two decades. Beyond clearing vegetation and reshaping valley
floors, it discharges **mercury, suspended sediment and turbidity into rivers**, with widely reported
consequences for water quality and for farmland downstream.

This research project **maps galamsey and studies its drivers and consequences**, in two phases:

1. **Remote sensing** — classifying mine presence from Landsat/Sentinel-2 imagery via Google Earth
   Engine (1995–2025, all Ghana).
2. **Spatial & economic analysis** — where mining occurs, why it clusters spatially, and what it does
   to the surrounding environment and economy.

The current analysis uses the **Barenblitt et al. (2021)** satellite-derived mining-extent maps for
**SW Ghana** (annual, 2007–2017) as an interim data source, on a **hexagonal grid** (~5 km cells) over
the study region.

**The specific piece this note supports** is an *event study* of the **downstream** effect of mining:
does galamsey **upstream** of a location degrade vegetation — and, by proxy, agricultural productivity
and water quality — **downstream**, through waterborne transport of mercury and sediment? We proxy the
downstream outcome with satellite **vegetation greenness (NDVI)**. To run that design we must know, for
every location, which mines are hydrologically **upstream** of it. That is entirely a question of **how
we define the river network and its direction of flow** — which is what this note is about, and where
we would most value a geologist's / hydrologist's input.

---

# The question we are unsure about {#intro}

This is a working note, not a settled write-up. We are trying to decide **what should count as a
"river"** for the galamsey study, and we would value a geologist's / hydrologist's view on whether we
are framing the problem sensibly at all.

"River" enters the analysis in **two different places**, and it is not obvious the same definition
should serve both:

1. **A proximity covariate — `dist_river_km`.** Distance to the nearest watercourse as a predictor of
   *where* galamsey occurs (alluvial gold sits in and beside channels). Here we need channel geometry.
2. **A directed flow graph — the event-study treatment.** The upstream→downstream design
   (`a_05_event_study.Rmd`) asks whether mining *upstream* of a hex degrades vegetation/water
   downstream. Here we need channels **and flow direction**, and the network must respect drainage
   divides.

Three candidate data sources encode "river" quite differently:

| Source | What a "river" is | Flow direction? | Small tributaries? |
|--------|-------------------|-----------------|--------------------|
| **OSM waterways** | A line a human mapped and labelled | No | Patchy — depends on mapping effort |
| **MERIT Hydro** | A DEM cell whose upstream drainage area exceeds a threshold | **Yes** (D8) | Yes, down to the chosen threshold |
| **Remote sensing** (MNDWI / JRC) | A pixel observed to hold surface water | No | Only where water is visible |

> **The worry that motivates this note.** In the waterborne event study, does the pollution signal
> from *large* rivers get diluted if we also label as "treated" many hexes that sit only on *tiny
> streams* — which may not actually transport mercury/sediment far? But if we raise the river
> threshold to keep only large channels, a large share of mining hexes fall out of the sample.
> **Is that the right way to think about the trade-off, and how should it be handled?** The section
> [*The event-study trade-off*](#tradeoff) is our attempt; we are genuinely unsure and would like your read.

We work on the **Ankobra basin** — the SW-Ghana river most associated with galamsey pollution, and the
area for which we already have a MERIT clip. Leaflet maps are used throughout because the
channel-vs-observed comparison is mostly visual (hence an `.Rmd`: leaflet embeds cleanly in a
self-contained `html_document`).

---

# Study area — the Ankobra corridor {#area}

```{r load-area}
# MERIT clip = the analysis window (the Ankobra corridor, native EPSG:4326).
hydro <- rast(merit_path)
names(hydro) <- c("dir", "upa", "wth", "elv")
dir_r <- hydro[["dir"]]; upa_r <- hydro[["upa"]]; wth_r <- hydro[["wth"]]; elv_r <- hydro[["elv"]]

merit_ext_ll  <- st_as_sf(as.polygons(ext(upa_r), crs = "EPSG:4326"))
merit_ext_utm <- st_transform(merit_ext_ll, UTM30N)

osm <- st_read(osm_path, quiet = TRUE) |> clean_names() |> st_make_valid() |>
  st_transform(4326)
osm_win <- suppressWarnings(st_crop(osm, st_bbox(merit_ext_ll)))

ankobra_river <- osm_win |>
  filter(str_detect(coalesce(name, ""), regex("ankobra", ignore_case = TRUE)))

gal <- st_read(barenblitt_path, quiet = TRUE) |> clean_names() |> st_make_valid() |>
  st_transform(4326)
mt_col <- intersect(c("mine_type", "minetype"), names(gal))[1]
gal_art <- gal[gal[[mt_col]] == 1, ]
gal_win <- suppressWarnings(st_crop(gal_art, st_bbox(merit_ext_ll)))

cat(sprintf("MERIT window: %.2f x %.2f deg | %d OSM segments | %d Ankobra-named | %d galamsey polys\n",
            diff(st_bbox(merit_ext_ll)[c(1,3)]), diff(st_bbox(merit_ext_ll)[c(2,4)]),
            nrow(osm_win), nrow(ankobra_river), nrow(gal_win)))
```

The MERIT clip is a narrow N–S strip: the Ankobra flows broadly south to the Gulf of Guinea. Galamsey
polygons (Barenblitt 2019 artisanal) cluster along the valley — which is, of course, part of what we
are trying to understand.

---

# Approach 1 — OSM waterways {#osm}

OpenStreetMap `waterway=*` lines are **crowd-sourced, human-labelled** features. They give real, named
rivers and a `waterway` **type** we can filter on. `d_03_waterways.R` keeps only *natural*
watercourses; artificial canals/drains/ditches presumably cannot host alluvial gold (is that a safe
assumption? — one for discussion).

```{r osm-classify}
natural    <- c("river", "stream", "brook", "wadi", "tidal_channel", "stream_pool", "flowline")
artificial <- c("canal", "drain", "ditch", "lock_gate", "fish_pass", "weir", "dam",
                "sluice_gate", "pressurised", "penstock")
osm_win <- osm_win |>
  mutate(ww_class = case_when(waterway %in% natural ~ "natural watercourse",
                              waterway %in% artificial ~ "artificial channel",
                              TRUE ~ "other / review"))

osm_win |> st_drop_geometry() |> count(waterway, ww_class, sort = TRUE) |>
  knitr::kable(caption = "OSM waterway types within the Ankobra window.")
```

```{r osm-leaflet, fig.height=7}
class_pal <- colorFactor(c(col_osm_nat, col_osm_art, col_osm_oth),
                         domain = c("natural watercourse", "artificial channel", "other / review"))
leaflet(osm_win) |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addProviderTiles("CartoDB.Positron", group = "Light") |>
  addPolylines(color = ~class_pal(ww_class), weight = 2, opacity = 1,
               label = ~paste0(coalesce(name, "(unnamed)"), " — ", waterway),
               group = "OSM waterways") |>
  addPolygons(data = gal_win, color = "white", weight = 1.2, fillColor = col_gal,
              fillOpacity = 0.65, group = "Galamsey (Barenblitt)") |>
  addLegend(pal = class_pal, values = ~ww_class, title = "OSM class", position = "bottomright") |>
  addLayersControl(baseGroups = c("Satellite", "Light"),
                   overlayGroups = c("OSM waterways", "Galamsey (Barenblitt)"),
                   options = layersControlOptions(collapsed = FALSE))
```

OSM gives us real named rivers and a type field, for free. But coverage seems to be a function of
*mapping effort* rather than hydrology — small tributaries (where alluvial galamsey may begin) look
unevenly mapped, there is no flow direction, and "river" here mixes trunk rivers with tiny mapped
streams while omitting unmapped ones. It may be fine as *a* proximity measure; it cannot build the
directed flow graph.

---

# Approach 2 — MERIT Hydro {#merit}

MERIT Hydro (Yamazaki et al. 2019, ~90 m) is a **DEM-derived, physically-modelled** river network,
with four bands used differently in this project:

| Band | Meaning | Use |
|------|---------|-----|
| `dir` | D8 flow direction (which neighbour each cell drains to) | The flow graph (upstream/downstream) |
| `upa` | Upstream drainage area (km²) accumulated through each cell | Defining channels via a threshold |
| `wth` | Satellite/OSM-fused river **width** (m); 0 = no mapped channel | A stricter, validated channel mask |
| `elv` | Hydrologically-adjusted elevation | HAND / vertical drop (not morphometry) |

## Why we use a `upa` threshold to define a "river" (and where we are unsure) {#upa-why}

`upa` is the **catchment area draining through each cell**, from flow accumulation over the D8 pointer.
Headwater cells have a tiny `upa`; moving downstream, tributaries merge and `upa` grows
**monotonically**. Our reasoning is that a river is *where concentrated flow persists*, so a cell is
"channel" when its `upa` exceeds a minimum catchment. On that logic **the threshold is the smallest
upslope area we are willing to call a channel:**

- **Low threshold** → a dense network reaching every gully (in the limit, every cell is a "stream").
- **High threshold** → only the major trunk rivers survive.

We have been treating the threshold as a modelling choice calibrated against OSM. **But is upstream
area even the right variable to threshold on?** A geologist might prefer discharge, stream order,
channel width, or a stream-power criterion — `upa` is a proxy for "how much water", not a measurement
of it. We would welcome a view on what the physically meaningful cut-off is for "this is a river that
can move sediment and contaminants".

```{r upa-dist, fig.height=3.6}
upa_vals <- values(upa_r, na.rm = TRUE)
qtab <- tibble(quantile = c("50%", "90%", "99%", "99.9%", "max"),
               upa_km2  = round(quantile(upa_vals, c(.5, .9, .99, .999, 1)), 2))
knitr::kable(qtab, caption = "Distribution of upstream drainage area (upa, km²) across Ankobra cells.")

tibble(upa = upa_vals[upa_vals > 0.05]) |>
  ggplot(aes(upa)) +
  geom_histogram(bins = 60, fill = "#2171B5") +
  geom_vline(xintercept = c(10, 50), colour = c("#E67E22", "#C0392B"), linetype = 2) +
  scale_x_log10(labels = comma) +
  annotate("text", x = 10, y = Inf, label = "ROUTE_KM2 = 10", vjust = 1.5, hjust = 1.05,
           colour = "#E67E22", size = 3) +
  annotate("text", x = 50, y = Inf, label = "STREAM_KM2 = 50", vjust = 3, hjust = -0.05,
           colour = "#C0392B", size = 3) +
  labs(title = "Where do the two working thresholds sit?",
       x = "Upstream drainage area (km², log scale)", y = "cells") +
  theme_minimal(base_size = 11)
```

## The channel network at several thresholds {#upa-thresholds}

```{r upa-networks}
THRESH <- c(5, 10, 20, 50, 500)
thr_summary <- map_dfr(THRESH, \(t) {
  n <- sum(upa_vals > t)
  tibble(`upa >` = paste0(t, " km²"), `channel cells` = n,
         `% of grid` = round(100 * n / length(upa_vals), 2))
})
knitr::kable(thr_summary,
             caption = "Channel-cell count by upa threshold. Lower cut = denser network reaching further up the tributaries.")
```

```{r upa-leaflet, fig.height=7}
thr_show <- c(10, 50)                       # ROUTE_KM2 vs STREAM_KM2
thr_cols <- c(col_upa10, col_upa50)
m <- leaflet() |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addProviderTiles("CartoDB.Positron", group = "Light")
for (i in seq_along(thr_show)) {
  mask_i <- ifel(upa_r > thr_show[i], 1L, NA)
  pal_i  <- colorNumeric(thr_cols[i], domain = c(0, 1), na.color = "transparent")
  m <- addRasterImage(m, mask_i, colors = pal_i, opacity = 0.95, project = TRUE,
                      group = sprintf("MERIT upa > %g km²", thr_show[i]))
}
m |>
  addPolylines(data = filter(osm_win, ww_class == "natural watercourse"),
               color = col_osm_nat, weight = 1.5, opacity = 0.9, group = "OSM natural") |>
  addPolygons(data = gal_win, color = "white", weight = 1.2, fillColor = col_gal,
              fillOpacity = 0.65, group = "Galamsey") |>
  addLegend(colors = c(col_upa10, col_upa50, col_osm_nat, col_gal),
            labels = c("MERIT upa > 10", "MERIT upa > 50", "OSM natural", "Galamsey"),
            title = "Layer", position = "bottomright") |>
  addLayersControl(baseGroups = c("Satellite", "Light"),
                   overlayGroups = c("MERIT upa > 10 km²", "MERIT upa > 50 km²",
                                     "OSM natural", "Galamsey"),
                   options = layersControlOptions(collapsed = FALSE)) |>
  hideGroup("MERIT upa > 50 km²")
```

Toggling the layers: at `upa > 50 km²` we get the trunk Ankobra and its main tributaries, which looks
close to the OSM *natural* "river" lines (this is roughly why `STREAM_KM2 = 50` was chosen). Dropping
to `upa > 10 km²` fills in a finer tributary network that OSM misses and that many galamsey polygons
sit on. **Which of these is the "real" river network for pollution transport is exactly what we are
unsure about.**

## Why two thresholds, and the tension between them {#two-thresholds}

The project currently uses **two** cuts for the two jobs from [the introduction](#intro):

- **`STREAM_KM2 = 50` — the "river" label.** Roughly where 4th-order tributaries begin; reproduces the
  OSM natural-river network.
- **`ROUTE_KM2 = 10` — the flow-graph routing cut.** Lower, so the directed graph reaches smaller
  tributaries. The stated justification (`d_04_merit.R` §11g): at `upa > 50` about **23% of mined
  hectares** fall off the channel network (never routed downstream), versus **~5.5% at `upa > 10`**.

The `d_04` §11j sweep over {2, 5, 10, 20} km² reports that the per-hex upstream-exposure *ranking* is
fairly stable across cuts, which we took to mean the exact value is not critical for the *routing*.
But that robustness is about *connectivity*, not about *how much pollution a channel can carry* — which
is the concern in the next section.

---

# The event-study trade-off: small streams vs. sample size {#tradeoff}

This is the heart of the note. In the waterborne design, a hex is (loosely) a candidate for
"waterborne treatment" if it sits on / near a channel that connects it to upstream mining. **The
channel threshold sets who is in that sample, and it cuts two ways:**

- **A low threshold (dense network).** Almost every mining hex counts as "on a river". But many of
  those channels are *tiny* — headwater streams with little discharge. If such streams do not
  actually transport mercury and sediment any meaningful distance, then we are labelling a lot of hexes
  as waterborne-connected when physically they are not. In regression terms this looks like
  **measurement error in the treatment**, which typically **attenuates** the estimated effect toward
  zero — the genuine signal from large, transport-capable rivers gets averaged together with many
  false-positive "treated" hexes.
- **A high threshold (sparse network).** The remaining channels are genuinely big rivers, so a
  "treated" hex is more plausibly *actually* exposed. But now a large share of mining hexes are **no
  longer on a river at all** and drop out of the waterborne sample. That costs power, and — more
  subtly — it may **change which mines are in the sample** (mines on big rivers may differ
  systematically from mines on small streams: size, accessibility, industrial vs artisanal), so it is
  not just a power loss but a possible **selection/composition** change.

**Is this the right framing?** We think the trade-off is real, but we are not sure "pick one
threshold" is the right response, for two reasons:

1. **The threshold conflates two things** — *connectivity* (is this hex hydrologically linked to a
   mine at all?) and *dose* (how much contaminant could actually arrive?). A binary on/off cut forces
   one number to do both jobs.
2. **Transport capacity is continuous, not a step.** Whether a stream moves galamsey pollution
   presumably rises smoothly with discharge / stream power, rather than switching on at 10 or 50 km².

Let us first *see* the trade-off in the Ankobra data, then list options — as questions, not answers.

## How many mining hexes are "at a river" at each threshold? {#elig-demo}

We take the 5 km hexes intersecting the Ankobra window, flag the **mining** hexes (Barenblitt
artisanal, `art_ha > 0`), and ask, for a grid of thresholds, whether each mining hex **contains a
MERIT channel cell** with `upa` above that threshold (i.e. does the mine sit on a channel that "counts"
as a river at that cut). This is the source side of the waterborne channel: a mine only injects
pollution into the routed network if it is on the network.

```{r elig-compute}
cs <- readRDS(hex_cache_path)
hex_sf <- cs$hex_sf
ha     <- cs$hex_analysis

hex_win <- hex_sf[lengths(st_intersects(hex_sf, merit_ext_utm)) > 0, ] |>
  left_join(select(ha, hex_id, art_ha, any_art), by = "hex_id")

# Max upstream area among each hex's MERIT cells → does it contain a channel > t?
hex_win$max_upa <- terra::extract(upa_r, vect(st_transform(hex_win, 4326)),
                                  fun = max, na.rm = TRUE)[, 2]
mine <- filter(hex_win, art_ha > 0)

THRG <- c(0, 5, 10, 20, 50, 100)
elig_tab <- map_dfr(THRG, \(t) tibble(
  `upa threshold (km²)`      = t,
  `mining hexes`             = nrow(mine),
  `at a river (eligible)`    = sum(mine$max_upa > t, na.rm = TRUE),
  `% retained`               = round(100 * mean(mine$max_upa > t, na.rm = TRUE), 1),
  `% dropped`                = round(100 * mean(!(mine$max_upa > t), na.rm = TRUE), 1)
))
knitr::kable(elig_tab, caption = paste0(
  "Ankobra window: of ", nrow(mine), " mining (artisanal) 5 km hexes, how many still sit on a ",
  "channel as the river threshold rises. Raising the cut from 10 to 50 km² drops ~1 in 4 mining hexes ",
  "out of the waterborne sample."))
```

```{r elig-curve, fig.height=4}
elig_tab |>
  ggplot(aes(`upa threshold (km²)`, `% retained`)) +
  geom_line(colour = "#2171B5", linewidth = 1) +
  geom_point(size = 2, colour = "#2171B5") +
  geom_vline(xintercept = c(10, 50), linetype = 2, colour = c("#E67E22", "#C0392B")) +
  scale_y_continuous(limits = c(0, 100)) +
  labs(title = "The sample-size cost of a stricter river threshold",
       subtitle = "Share of Ankobra mining hexes that remain 'on a river' as the upa cut rises",
       x = "upa threshold (km²)", y = "% of mining hexes retained") +
  theme_minimal(base_size = 11)
```

```{r elig-riversize, fig.height=4}
# How big is the channel each mining hex sits on? This is the flip side of the worry:
# how many mining hexes are only on small streams?
pct_small10 <- round(100 * mean(mine$max_upa <= 10, na.rm = TRUE))
pct_small50 <- round(100 * mean(mine$max_upa <= 50, na.rm = TRUE))
mine |> st_drop_geometry() |>
  ggplot(aes(pmax(max_upa, 0.1))) +
  geom_histogram(bins = 30, fill = "#8856A7") +
  geom_vline(xintercept = c(10, 50), linetype = 2, colour = c("#E67E22", "#C0392B")) +
  scale_x_log10(labels = comma) +
  labs(title = "What size river does each mining hex sit on?",
       subtitle = sprintf("~%d%% of mining hexes sit on a channel ≤ 10 km²; ~%d%% ≤ 50 km² (upstream area).",
                          pct_small10, pct_small50),
       x = "Largest channel in the hex — upstream area (km², log scale)",
       y = "mining hexes") +
  theme_minimal(base_size = 11)
```

```{r elig-facetmap, fig.height=7}
# Where are the eligible vs dropped mining hexes, as the threshold rises?
thr_facets <- c(5, 10, 20, 50)
osm_nat_ll <- filter(osm_win, ww_class == "natural watercourse")
elig_long <- map_dfr(thr_facets, \(t) hex_win |>
  st_transform(4326) |>
  mutate(threshold = factor(paste0("upa > ", t, " km²"),
                            levels = paste0("upa > ", thr_facets, " km²")),
         status = case_when(art_ha > 0 & max_upa > t ~ "mine — on a river (kept)",
                            art_ha > 0               ~ "mine — off network (dropped)",
                            TRUE                     ~ "no mine")))

ggplot(elig_long) +
  geom_sf(aes(fill = status), colour = "grey85", linewidth = 0.1) +
  geom_sf(data = osm_nat_ll, colour = "#3B8ED0", linewidth = 0.3, inherit.aes = FALSE) +
  facet_wrap(~ threshold, nrow = 1) +
  scale_fill_manual(values = c("mine — on a river (kept)"    = "#1A9850",
                               "mine — off network (dropped)" = "#D73027",
                               "no mine"                      = "grey92"),
                    name = NULL) +
  labs(title = "Which mining hexes stay in the waterborne sample as the river threshold rises",
       subtitle = "Ankobra window, 5 km hexes. Green = mine on a channel at that cut; red = mine dropped; blue = OSM natural rivers.",
       caption = "Barenblitt artisanal mining; MERIT Hydro upa. Higher threshold = more red (mines fall out).") +
  theme_void(base_size = 10) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"),
        plot.caption = element_text(colour = "grey50", size = 7))
```

The red hexes are the cost of a stricter definition; the green are the "confidently on a real river"
sample. The histogram is the other side of the coin: it shows how many mines sit only on small
channels in the first place.

## Ways this might be addressed — options and questions {#options}

We are unsure which of these is right (or whether the premise holds). Candidates, roughly from least
to most structural:

1. **Treat the threshold as a robustness axis, not a single choice.** Estimate the downstream effect
   across the whole grid of cuts and report the path. *If* the effect **strengthens** as we restrict to
   larger rivers, that is arguably *evidence for* the waterborne mechanism (transport capacity
   matters), not a nuisance. If it is flat, the small-stream dilution worry may be overstated.
   **Question:** would you expect the effect to grow with river size if the mechanism is real?
2. **Use a continuous, flow-weighted dose instead of a binary "on a river".** Weight each mine's
   contribution by the discharge / `upa` / channel width `wth` at the connecting reach, so a mine on
   the trunk Ankobra counts for more than one on a trickle. This keeps every mine in the sample
   (no drop-outs) while down-weighting small streams. (`d_04` already prototypes a distance-decayed
   upstream exposure — this would extend it with a flow weight.) **Question:** is `upa` an acceptable
   proxy for transport capacity, or is discharge/stream power needed?
3. **Stratify by stream order / size class and estimate separately.** Headwater vs mid-order vs trunk.
   This tests the mechanism directly rather than assuming a cut-off. MERIT can yield Strahler order
   (`d_04` §10d). **Question:** is Strahler order the right stratifier, or something else?
4. **Interact treatment with a flow measure.** Keep the low threshold (retain the sample) but interact
   upstream-mining treatment with the receiving reach's `upa`/`wth`. If the effect loads on the
   high-flow interaction, the dilution worry is *quantified* rather than assumed.
5. **A physically-motivated decay/dilution kernel.** Model contaminant concentration falling with
   downstream distance, with a decay length that depends on discharge — so small streams "transport"
   only a short way by construction. **Question:** does a simple dilution/settling model exist that we
   could parameterise for mercury vs. suspended sediment?
6. **Separate the two roles of the threshold.** Use a low cut for *connectivity* (don't drop mines
   from the graph) and a flow-based *dose* for intensity. This directly answers the two-jobs-one-number
   problem in [the trade-off section](#tradeoff).

### Open questions for a geologist / hydrologist {#geo-questions}

These are the questions where we most need expertise, because they determine whether the trade-off
above is even the binding one:

- **How far does galamsey pollution actually travel, and does that depend on stream size?** Is mercury
  mostly **sediment-bound** (settles quickly → proximity/large-river matters) or **dissolved /
  methylated** (travels far → even small streams connect)? These imply very different "right"
  thresholds.
- **Is there a physical catchment/discharge size below which a mine's load is simply not detectable
  downstream?** If so, that number — not a statistical convenience — should set the threshold.
- **Do turbidity/sediment and mercury have different transport lengths?** We use NDVI as the outcome,
  which may respond to sediment/turbidity more than to mercury — should the "river" definition differ
  by which pollutant/outcome we care about?
- **Is upstream drainage area an acceptable channel proxy at ~90 m, or should we be using channel
  width (`wth`), modelled discharge, or stream power instead?**
- **Are dry-season vs. wet-season flows so different that an annual "is this a river" label is
  meaningless for parts of the network?**

---

# The other MERIT bands for the Ankobra {#merit-bands}

For completeness, the bands beyond the threshold discussion.

### `upa` — upstream drainage area

```{r upa-map, fig.height=6}
plot(log1p(upa_r), col = hcl.colors(50, "Blues", rev = TRUE),
     main = "log(1 + upstream drainage area) [km²] — Ankobra")
```

`upa` rises smoothly downstream; the bright trunk is the Ankobra itself (max ≈ 8,400 km² at the
outlet). Besides thresholding, its log is a natural **"downstreamness" scalar** (higher = further down
the network).

### `wth` — river width {#wth}

```{r wth-map, fig.height=6}
wth_ch <- wth_r; wth_ch[wth_ch == 0] <- NA
plot(wth_ch, col = hcl.colors(30, "Blues", rev = TRUE),
     main = "MERIT river channel width (m) — Ankobra (0 = no mapped channel)")
```

```{r wth-summary}
wth_pos <- values(wth_r, na.rm = TRUE); wth_pos <- wth_pos[wth_pos > 0]
cat(sprintf("wth non-zero cells: %d (%.2f%% of grid) | median %.1f m | max %.1f m\n",
            length(wth_pos), 100 * length(wth_pos) / length(upa_vals),
            median(wth_pos), max(wth_pos)))
```

`wth` is a **satellite + OSM-fused, validated river width**, non-zero on only a small fraction of
cells (the confidently-mapped channels). It might be a more *physical* channel definition than a `upa`
threshold — it marks reaches observed to be rivers — but it will miss narrow headwater tributaries. It
is tempting as a middle ground; whether it is better than a `upa` cut for our purpose is unclear to us.

```{r wth-leaflet, fig.height=7}
pal_wth <- colorNumeric("plasma", domain = wth_pos, na.color = "transparent")
leaflet() |>
  addProviderTiles("Esri.WorldImagery", group = "Satellite") |>
  addProviderTiles("CartoDB.Positron", group = "Light") |>
  addRasterImage(ifel(upa_r > 10, 1L, NA),
                 colors = colorNumeric(col_upa10, c(0, 1), na.color = "transparent"),
                 opacity = 0.6, project = TRUE, group = "MERIT upa > 10 km²") |>
  addRasterImage(wth_ch, colors = pal_wth, opacity = 1, project = TRUE,
                 group = "MERIT width (wth)") |>
  addPolylines(data = filter(osm_win, ww_class == "natural watercourse"),
               color = col_osm_nat, weight = 1.5, opacity = 0.9, group = "OSM natural") |>
  addLegend(pal = pal_wth, values = wth_pos, title = "wth (m)", position = "bottomright") |>
  addLayersControl(baseGroups = c("Satellite", "Light"),
                   overlayGroups = c("MERIT width (wth)", "MERIT upa > 10 km²", "OSM natural"),
                   options = layersControlOptions(collapsed = FALSE))
```

### `elv` — hydrologically-adjusted elevation

```{r elv-map, fig.height=6}
plot(elv_r, col = hcl.colors(50, "Terrain2"),
     main = "MERIT hydro-adjusted elevation (m) — Ankobra")
```

`elv` is nudged cell-by-cell to enforce downhill flow — right for HAND (vertical drop to nearest
channel), but not for slope/morphometry (use MERIT-**DEM** there).

---

# Approach 3 — remotely sensed surface water {#rs}

A third notion of "river" is **observed surface water** — pixels that actually hold water. Two
flavours: (1) our own **MNDWI** from Landsat/Sentinel-2 (`rs01`/`rs02`), which with NDTI can even flag
*turbid* mining ponds; and (2) **JRC Global Surface Water** (`JRC/GSW1_4/GlobalSurfaceWater`), a
validated occurrence layer (1984–2021, 30 m). Neither is downloaded yet, so the code below is a
template (needs GEE auth) and is not evaluated on knit.

```{r rs-jrc-template, eval=FALSE}
library(rgee)
bb  <- st_bbox(merit_ext_ll)
aoi <- ee$Geometry$Rectangle(c(bb[["xmin"]], bb[["ymin"]], bb[["xmax"]], bb[["ymax"]]))
occurrence <- ee$Image("JRC/GSW1_4/GlobalSurfaceWater")$select("occurrence")$clip(aoi)
mndwi <- ee$ImageCollection("COPERNICUS/S2_SR_HARMONIZED")$
  filterBounds(aoi)$filterDate("2023-11-01", "2024-03-01")$
  filter(ee$Filter$lt("CLOUDY_PIXEL_PERCENTAGE", 40))$median()$
  normalizedDifference(c("B3", "B11"))$rename("mndwi")$clip(aoi)
# Export to Drive, read back with terra::rast(), overlay in leaflet as above.
```

This observes *actual* water, including mining ponds the other sources never encode — but it detects
*water bodies*, not a connected channel graph; it is seasonal and cloud-limited; narrow rivers under
canopy are invisible; and it gives no flow direction. It seems best as a **complement** (validating
MERIT channels, flagging ponds) rather than the definition of the network.

---

# Does the definition move `dist_river_km`? {#compare}

A practical check: on the 5 km hexes inside the Ankobra window, how much does distance-to-nearest-river
shift across definitions? (Mirrors `d_03_waterways.R` §5–7, restricted to the MERIT extent.)

```{r dist-compare, fig.height=5}
cent    <- st_centroid(hex_sf)
inside  <- lengths(st_intersects(cent, merit_ext_utm)) > 0
hex_in  <- cent[inside, ]

osm_all_utm <- st_transform(osm_win, UTM30N)
osm_nat_utm <- st_transform(filter(osm_win, ww_class == "natural watercourse"), UTM30N)
chan_pts <- function(thr) {
  m <- ifel(upa_r > thr, 1L, NA)
  suppressWarnings(as.points(m) |> st_as_sf() |> st_transform(UTM30N))
}
merit10 <- chan_pts(10); merit50 <- chan_pts(50)
wth_pts <- suppressWarnings(as.points(ifel(wth_r > 0, 1L, NA)) |> st_as_sf() |>
                              st_transform(UTM30N))

dist_to <- function(pts, tgt) {
  idx <- st_nearest_feature(pts, tgt)
  as.numeric(st_distance(pts, tgt[idx, ], by_element = TRUE)) / 1000
}
dcmp <- tibble(
  hex_id          = hex_sf$hex_id[inside],
  `OSM all`       = dist_to(hex_in, osm_all_utm),
  `OSM natural`   = dist_to(hex_in, osm_nat_utm),
  `MERIT upa>50`  = dist_to(hex_in, merit50),
  `MERIT upa>10`  = dist_to(hex_in, merit10),
  `MERIT wth>0`   = dist_to(hex_in, wth_pts)
)
dcmp |>
  pivot_longer(-hex_id, names_to = "definition", values_to = "dist_km") |>
  group_by(definition) |>
  summarise(mean = round(mean(dist_km), 2), median = round(median(dist_km), 2),
            max = round(max(dist_km), 2), .groups = "drop") |>
  knitr::kable(caption = "Per-hex distance to nearest river (km) by definition — Ankobra window.")
```

```{r dist-scatter, fig.height=4.2}
dcmp |>
  pivot_longer(c(`MERIT upa>50`, `MERIT upa>10`, `MERIT wth>0`),
               names_to = "merit_def", values_to = "merit_km") |>
  ggplot(aes(`OSM natural`, merit_km)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60", linetype = 2) +
  geom_point(alpha = 0.4, size = 0.7, colour = "#2171B5") +
  facet_wrap(~ merit_def) + coord_equal() +
  labs(title = "Distance to river: OSM natural vs MERIT definitions (5 km hexes)",
       subtitle = "Below the 1:1 line = MERIT finds a nearer channel than OSM",
       x = "OSM natural (km)", y = "MERIT (km)") +
  theme_minimal(base_size = 10)
```

The denser MERIT network (`upa > 10`) places almost every hex close to a channel; `wth` / `upa > 50`
behave more like the OSM natural rivers. Which is "better" again depends on the use, and on the
mechanism questions in [*Open questions for a geologist*](#geo-questions).

---

# Where we have landed — provisionally {#synthesis}

Our *current, tentative* leaning (open to revision):

| Use | Leaning | Why — and our uncertainty |
|-----|---------|---------------------------|
| **Flow graph / upstream treatment** | MERIT `dir` + a **low** `upa` cut (10 km²) | Only source with flow direction; the low cut keeps small-tributary mines in the graph. *But* whether those small channels should carry treatment weight is exactly [the trade-off](#tradeoff). |
| **`dist_river_km` covariate** | MERIT `upa > 50` or OSM natural (they agree) | Reproduces the observed river network; robustness = swap between them. |
| **Pollution-source / pond validation** | Remote sensing (MNDWI, JRC) | Sees real standing water incl. galamsey ponds — a complement, not the network. |

The part we are **least** sure about is the treatment-definition trade-off ([above](#tradeoff)): whether
to pick a threshold, sweep it, weight by flow, or stratify by stream size. Our instinct is that a
**continuous flow-weighted dose** (option 2/4) is more honest than any single cut — but that rests on
`upa`/`wth` being an acceptable proxy for how far galamsey pollution actually travels, which is a
question for someone who knows the geochemistry. **Feedback on the [open questions](#geo-questions) would
change what we build next.**

---

*Data: MERIT Hydro (Yamazaki et al. 2019); OpenStreetMap contributors; Barenblitt et al. (2021).
Companion scripts: `d_03_waterways.R` (OSM), `d_04_merit.R` (MERIT flow graph), `d_05_ndvi.R` (Ankobra
basin definition). See also `4_presentation/methodology_explainer.md` (event-study designs).*
